---
title: "TimeToEventData"
output: html_document
date: "2024-05-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background
In the paper, "" we describe how to quantify repeatability from time-to-event data. In this supplementary materials page we demonstrate the use of this method with real data in 3 worked examples. Secondly, we present the methods and results for a simulation study to illustrate the relationship among ICC (repeatability) values derived from 4 different models and residual variance estimators.

## Worked Examples
Behavior data range widely in the number of individuals as well as the number of repeated measures on each individuals. Furthermore, the presence and amount of censored data can vary. We use openly available data to present 3 worked examples that 

### Example 1: Frog aggression towards a simulated intruder
Our first example data set is from Peignier et al. 2022. They investigated whether *A. femoralis* frogs show repeatable aggressive behavior by measuring the latency of territorial males to approach a simulated intruder. There sample size included 51 males with 3.2 +/- 1.31 trials each. Censored data are present (6% of trials) because some males did not approach the simulated intruder. These data were given a ceiling value of 300 seconds (the longest trial time). The authors log-transformed this variable and used the Gaussian family in the rpt function (Stoffel et al. 2017).
```{r E1 - load the data, echo=FALSE}
library(coxme)
library(survival)

# Data from Peignier et al. 2022

fdata = read.csv("Frog_personality.csv")
fdata = fdata[which(fdata$sex == "m"),c(1,2,9)] # Only males participated in the aggression test, we only need columns for ID, repetition and latency to approach the simulated intruder. Authors did not include fixed effects, so we won't either.
fdata = fdata[!is.na(fdata$jump_s),] # Unclear what the NAs represent, but the authors remove these data, so we will too.
fdata$event = ifelse(fdata$jump_s==300, 0,1) # Create event column that indicates whether the data were censored for that trial.

head(fdata,20)
```
This is a classic (un-exploded) format for animal personality data. However, to get the variance components for our time-to-event ICC equation, we need a data set that includes discrete time intervals. This is referred to as an "exploded" data set and we can use the function survSplit in the survival package (cite) to easily transform the data. 

How to choose interval size... (fill in based on simulation results)

```{r E1 - explode the data, echo=F}
# decide on the intervals
psych::describe(fdata$jump_s)

fdata_int = survSplit(Surv(jump_s,event) ~ repetition+ID, data = fdata, 
                      cut = c(50,100,150,200,250), start = "tstart",end = "tstop")

head(fdata_int,20)
```

Running the cox proportional hazards model

```{r E1 - cox model and ICC calculation, echo = F}
c1 = coxme(Surv(tstart,tstop,event)~1+(1|ID), data = fdata_int)

var <- VarCorr(c1)$ID[[1]]
fr.lognormal(k,s,var,what = "tau")
var/(var + pi^2/6)
# 0.19
# Original paper reports a repeatabiltiy of 0.24 (CI: 0.07-0.40)

```


### Example 2: Christmas tree worm re-emergence after a simulated predator attack
These data come from Pezner et al. 2017 https://academic.oup.com/beheco/article/28/1/154/2453511. They tested the consistency of hiding time (latency to re-emerge from the hole) within and across days, and the impact of social environment. They tested 30 worms, each received 4 trials within a day and 4 days of sampling for a total of 16 trials per worm. Censored data are not explicitly stated, but there are 5 NA values from 2 individuals that we here assume are censored data. The authors log-transformed the hiding time variable, used a linear mixed model to get variance estimates, then hand-calculated repeatability.

```{r E2, echo=F}
## Load the data
ctw = read.csv("CTWemergence.csv")
# "HT" variable indicates hiding time, or the latency to emerge; "Whorls" is a visual indicator of age

# 2 individuals have NA values, but it is not explained why. 
# I'll assume these are censored (the worm didn't emerge in the trial time) and give ceiling value of one unit after other highest value
ctw$event = ifelse(is.na(ctw$HT),0,1)
ctw$HT[which(is.na(ctw$HT))]<- 375

## Explode the data
# decide on the intervals
psych::describe(ctw$HT)
ctw_int = survSplit(Surv(HT,event) ~ Whorls + Trial_Total + Worm_ID, data = ctw, 
                    cut = c(50,100,150,200,250,300,350,400), start = "tstart",end = "tstop")

## Cox model and ICC calculation
ctw.cox.int = coxme(Surv(tstart,tstop, event)~Whorls + (1|Worm_ID), data=ctw_int)

var <- VarCorr(ctw.cox.int)$Worm_ID[[1]]
var/(var + pi^2/6) #0.33
# Original paper finds across day hiding time repeatability is 0.42
```



### Example 3: Repeatability of distance to cache a seed in smalll mammals 
Brehm et al. 2019 (https://onlinelibrary.wiley.com/doi/full/10.1111/ele.13324) evaluated seed caching behavior in relation to multiple personality traits and habitat characteristics. They offered artificial seeds for small mammals to cache that had reflective flags attached so that the cached seeds could be relocated and the distance from the food platform could be measured. The data include 31 unique deer mice that cached between 1-12 seeds (mean = 3.31 &pm; 0.5). Understandably, it can be hard to relocate the cached seeds and so 35% of the caching distances were censored (seeds lost to follow up).

```{r E3, echo = F}
## Load the data
data<-read.csv(file="data_seed_pers.csv")
pm<-subset(data, SPP=="PM") # only one species

pmdistmov<-subset(pm,REMOVE==1) # distance the seed is dispersed, only looking at seeds that were removed from the feeding platform
pmdistmov<-subset(pmdistmov, CONS!=1) # remove rows where the seed was consumed close by the feeding platform
pmdistmov$RECOVERED..Y.N.[which(pmdistmov$DIST..MOVED==15)]<-"Y" # one row seems to indicate the cached seed was found 15m from the feeding platform, but the categorical variable for whether it was removed is an NA 
table(pmdistmov$RECOVERED..Y.N.) # how many cached seeds did they recover (i.e., how much censored data is there?)
# 35% of data are censored
dist = pmdistmov[-which(pmdistmov$ID == "UNK"),c(1,5,6,14,17)] #simplify data frame
# the Recovered column is analogous to the event variable in surivival analysis. Modify it to be an integer
dist$event = ifelse(dist$RECOVERED..Y.N.== "Y",1,0)

## Explode the data
# Create intervals based on the number of events in each interval. Here, 25% of the data are in each of 4 intervals.
quantile(dist$DIST..MOVED, probs = seq(0,1,0.25),na.rm=T)

psych::describe(dist$DIST..MOVED)
dist$DIST..MOVED[which(is.na(dist$DIST..MOVED))]<-1038 # give ceiling value to NAs

dist_int = survSplit(Surv(DIST..MOVED, event) ~ ID, data = dist, 
                     cut = c(50,146,331.5), start = "tstart",end = "tstop")

## Cox model and ICC calculation
dist.cox.int = coxme(Surv(tstart,tstop, event) ~ 1 + (1|ID), data=dist_int)
var <- VarCorr(dist.cox.int)$ID[[1]] 
var/(var + pi^2/6) # 0.05
# Original paper did not run this particular analysis.
```



## Simulations



